---
title: "Heart Disease UCI Analysis"
author: "Kevin Tran"
date: "9/7/2020"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls()) #removes all variables stored previously in Environment (good habit to include)

library(tidyverse) #Run tidyverse, if you do not have: install.packages("tidyverse")
library(forcats) # transform integers to factors

#read the data set into RStudio and stored into object
data <- read.csv("C:/Users/Kevin/Desktop/Heart-Disease-UCI/datasets_33180_43520_heart.csv")

heart_data <- as_tibble(data)

heart_data <- rename(heart_data, age = ï..age)

heart_data <- heart_data %>%
  mutate(status = target)

heart_data$sex[heart_data$sex == "1"] <- "male" #change values from 1 to male in the sex variable
heart_data$sex[heart_data$sex == "0"] <- "female" #change values from 0 to female in the sex variable
heart_data$sex <- as.factor(heart_data$sex)

heart_data$cp <- as.factor(heart_data$cp)
heart_data$fbs <- as.factor(heart_data$fbs)
heart_data$restecg <- as.factor(heart_data$restecg)
heart_data$exang <- as.factor(heart_data$exang)
heart_data$slope <- as.factor(heart_data$slope)

heart_data$status[heart_data$status == "0"] <- "Dz" #change values from 0 to disease in the target variable
heart_data$status[heart_data$status == "1"] <- "No Dz" #change values from 1 to no disease in the target variable
heart_data$status <- as.factor(heart_data$status)

tidy_data <- heart_data %>% # main data set used throughout the data analysis
  select(age:oldpeak, status) 

```

## Abstract

WORK ON AFTER CODE CHUNKS AND ORGANIZATION IS FINALIZED

## Sex and Disease Status

We will begin our analysis by determining the prevalence of each respective sex within the data set.

```{r}
ggplot(tidy_data, mapping = aes(x = sex)) +
  geom_bar()
```

* As shown in the above bar chart, there are more males than females within this data set.

**Let's take a look at the relationship between sex (male vs female) and disease status:**
```{r}
ggplot(tidy_data, mapping = aes(x = sex, fill = status)) +
  geom_bar(position = "fill")
```

* Within the females of this analysis, roughly 25% have heart disease and within the males of this analysis, roughly 60% have heart disease.

According to *Harvard Health*, "At younger ages, men face a greater risk of heart disease than women." On average, men will experience their first heart attack at 65 years old and women will experience their first heart attack at 72 years old.
<https://www.health.harvard.edu/heart-health/the-heart-attack-gender-gap#:~:text=At%20younger%20ages%2C%20men%20face,first%20heart%20attack%20is%2072.>

*Therefore, the above proportions are not out of the ordinary since males are more susceptible to heart disease earlier in life than females.*

## Age and Disease Status

According to the *National Institute on Aging*, adults that are 65 years and older have a higher risk of suffering heart complications. Therefore, we will analyze the data and see what the average age of individuals with and without heart disease is within our data set.

```{r, message=FALSE}
ggplot(tidy_data, mapping = aes( x = status, y = age)) +
  geom_boxplot()

average_age_dz <- tidy_data %>%
  group_by(status) %>%
  summarise(
    average_age = mean(age),
    min = min(age),
    max = max(age)
  )

knitr::kable(average_age_dz)
```

* Within our data set, the average age of an individual with heart disease is 56.6 years old with the youngest being 35 years old and the oldest being 77 years old.

* The box plot above shows that the average age of an individual with heart disease is greater than that of someone without heart disease and there is less variability in age of those with heart disease than those without heart disease.

**Now we can run a t-test keeping in mind, our null hypothesis is equal to zero, the alternative is two.sided, we want a 95% confidence interval, the variance is not equal and we have independent populations.**
```{r}
t.test(tidy_data$age ~ tidy_data$status, mu = 0, alt = "two.sided", conf = 0.95, var.eq = FALSE, paired = FALSE)
```
* The null hypothesis is: there is no difference in average age between those with or without heart disease.

* The p-value is close to zero, therefore we can reject our null hypothesis and with 95% confidence say there is a statistically significant difference between the average age of those who have heart disease vs those who do not.

* The confidence interval also does not include zero, therefore, it supports our rejection of the null.

As mentioned in the previous section, heart disease status is influenced (not limited to) by both age and sex (male vs female). Let's explore the average age of males vs females with and without heart disease and compare it to the claim from, *Kaiserpermanente.org*, that "men tend to develop coronary artery disease earlier in life."
<https://wa.kaiserpermanente.org/healthAndWellness/index.jhtml?item=%2Fcommon%2FhealthAndWellness%2Fconditions%2FheartDisease%2FageAndGender.html>

It is also known that heart disease "develops 7-10 years later in women than in men." According to the article, <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3018605/#:~:text=Women%2C%20Risk%20Factors-,Cardiovascular%20disease%20develops%207%20to%2010%20years%20later%20in%20women,'protected'%20against%20cardiovascular%20disease.>

```{r, message = FALSE}
age_sex_dz <- tidy_data %>%
  group_by(status, sex) %>%
  summarise(
            average_age = mean(age),
            min = min(age),
            max = max(age)
            ) %>%
  filter(status == "Dz")

knitr::kable(age_sex_dz)
```
Within this analysis you will see the following:

* Average age of males with heart disease is 56.1 years old and the average age of females with heart disease is 59 years old.

* The youngest male with heart disease is 35 years old and the youngest female with heart disease is 43 years old.

* The oldest male with heart disease is 77 years old and the oldest female with heart disease is 66 years old.

This analysis is aligned with the claims stated above. The average age of males with heart disease was on average 3 years less than that of a female which supports the claim that "men tend to develop coronary artery disease earlier in life." If we look at the youngest male to develop heart disease, they are 8 years younger than the youngest female with heart disease in this data set. Which supports the claim that heart disease "develops 7-10 years later in women than in men."

## Cholestoral Levels and Disease Status

*Webmd.com*, states that high cholesterol is asymptomatic, therefore, many people are unaware of their high cholesterol levels. However, if too much cholesterol is built up in the walls of your arteries, the arteries become narrow and blood flow to the heart becomes slowed down or blocked. This can result in reduced oxygen to the heart and result in chest pain. If complete blockage occurs, this may result in a heart attack.

The bad cholesterol that causes blockage of the arteries is called Low-Density Lipoprotein (LDL) and the good cholesterol which works to clear bad cholesterol from the blood is called High-Density Lipoprotein (HDL).
<https://www.webmd.com/heart-disease/guide/heart-disease-lower-cholesterol-risk#1>

**We will take a look at our data set and see what the serum cholesterol levels are compared to the normal range of serum cholesterol:**
```{r, message=FALSE, error=FALSE}
tidy_data %>%
  group_by(status, chol) %>%
  ggplot(tidy_data, mapping = aes(x = status, y = chol)) +
  geom_boxplot()

average_chol_status <- tidy_data %>%
  group_by(status) %>%
  summarise(
            average_chol = mean(chol),
            min = min(chol),
            max = max(chol)
            )

knitr::kable(average_chol_status)
```

As shown above, the average serum cholesterol levels of those with heart disease was 251 mg/dL and the average serum cholesterol levels of those without heart disease was 242 mg/dL.

According to *WebMD*:

* The desirable range of serum cholesterol level is less than 200mg/dL.

* The borderline high of serum cholesterol level is: 200-239 mg/dL

* The High of serum cholesterol level is: greater than 240 mg/dL (being at this stage may double the risk of heart disease)

Therefore, both of the average serum cholesterol levels of those with and without heart disease fall within the "High of Serum Cholesterol Level."
<https://www.webmd.com/cholesterol-management/understanding-your-cholesterol-report>

**Although, they are both above the health range, are the two averages significantly different?**
```{r}
t.test(tidy_data$chol ~ tidy_data$status, mu = 0, alt = "two.sided", conf = 0.95, var.eq = FALSE, paired = FALSE)
```
* Our null hypothesis states: There is no difference between the average serum cholesterol levels of a patient with heart disease and of a patient without heart disease.

* Our p-value shows that the probability of getting our results/extremes, assuming the null hypothesis is true, will happen 13.6% of the time. Therefore, we cannot reject our null hypothesis.

* Also since the 95% confidence interval includes zero, it supports our failure to reject the null hypothesis.

**Therefore, there is no statistically significant difference between the two cholesterol levels.**

It is interesting to see that cholesterol level is not a good indicator of heart disease within this data set despite the known implications from *WebMd.com*.

## Chest Pain and Disease Status

Chest pain, also known as Angina, is not a disease but a symptom of a potential underlying heart problems. There are many different variations of Angina, however, we will be following the below factors for our analysis:

```{r}
ggplot(tidy_data, aes(x = cp, fill = status)) +
  geom_bar(position = "stack") +
  xlab("Chest Pain Type")
```

* Value 0: Asymptomatic

* Value 1: Atypical Angina

* Value 2: Non-Anginal Pain

* Value 3: Typical Angina

As shown in the above plot, the majority of patients within the data set are asymptomatic. Within this majority, most of which have heart disease. This indicates that the majority of patients with heart disease are asymptomatic. 

According to *heart.org*, "Not all chest pain is a sign of heart disease." There are other conditions that can cause chest pains, such as, "Pulmonary Embolism, Aortic Dissection, Lung infection, Aortic Stenosis, and etc.

**Therefore, as seen above, we can assume that chest pain might not be a great indicator for heart disease as well.**

## Resting Blood Pressure and Disease Status

As stated in *MedicalNewsToday*, *"when doctors evaluate the risk of high blood pressure, they usually pay more attention to systolic blood pressure, which they consider a major risk factor for cardiovascular disease in older adults."* They also mention that high diastolic blood pressure is also indicative of cardiovascular disease, however, our data set will only include systolic blood pressure.

* We will focus on the normal ranges of systolic blood pressure since diastolic blood pressure is out of the scope of this data analysis.

The normal ranges of systolic blood pressure is less than 120 mm Hg. Systolic blood pressure that is consistently greater than 130 mmHg categorizes the patient as hypertensive.

<https://www.medicalnewstoday.com/articles/325861>

**Below is a Histogram to show the resting systolic blood pressure of patients with and without heart disease:**

```{r, message=FALSE}
ggplot(tidy_data, aes(x = trestbps, fill = status)) +
  geom_histogram() +
  xlab("Resting Blood Pressure (mmHg)")
```

* As you can see, the histogram is positively skewed due to the outliers of high resting blood pressure greater than 180 mmHg (which both patients suffer from heart disease).

```{r, message=FALSE}
trestbps_table <- tidy_data %>%
  group_by(status) %>%
  summarise(
    mean(trestbps)
  )

knitr::kable(trestbps_table)
```

* As calculated above, the average resting blood pressure of patients with heart disease is 134 mmHg (which is outside of the healthy range) and could result in being hypertensive (depending on consistent readings of >130 mmHg).

* The average resting blood pressure of patients without heart disease is 129 mmHg, which is still greater than the healthy range but is less than a hypertensive patient.

**Can we claim that the average blood pressures between those with or without the heart disease is significantly different?**

```{r}
t.test(tidy_data$trestbps ~ tidy_data$status, mu = 0, alt = "two.sided", conf = 0.95, var.eq = FALSE, paired = FALSE)
```

* The null hypothesis is: there is no difference in average systolic blood pressure between those with or without heart disease.

* The p-value is 1.3% which is less than 5%, therefore we can reject our null hypothesis and with 95% confidence say there is a statistically significant difference between the average systolic blood pressure of those who have heart disease vs those who do not.

* The confidence interval also does not include zero, therefore, it supports our rejection of the null.


## Maximum Heart Rate Achieved per Patient and Disease Status

```{r, message=FALSE}
ggplot(tidy_data, aes(x = thalach, fill = status)) +
  geom_histogram() +
  xlab("Maximum Heart Rate Achieved")
```

Upon researching the average maximum heart rate, *mayoclinic.com* <https://www.mayoclinic.org/healthy-lifestyle/fitness/in-depth/exercise-intensity/art-20046887#:~:text=You%20can%20calculate%20your%20maximum,beat%20per%20minute%20during%20exercise.> suggests that it varies by age. To calculate the average maximum heart rate: you will subtract your age from 220.

* This implies that the older you are the lower your maximum heart rate is (on average).
Therefore, we will look at a plot that includes age as a variable and see what it implies.

```{r, message = FALSE}
ggplot(tidy_data, aes(x = age, y = thalach, color = status)) +
  geom_point() +
  geom_smooth(se = FALSE, size = 2) +
  xlab("Age") +
  ylab("Maximum Heart Rate Achieved")

```

* As seen in the above graph, this is true for those without heart disease. As age increases, the maximum heart rate achieved decreases.

* Within patients with heart disease the maximum heart rate achieved remains consistent around 140-150 beats per minute regardless of age.

**We will now look at a linear model to confirm relationship between heart rate, age, and disease status:**
```{r}
thalach_age_model <- lm(tidy_data$thalach ~ tidy_data$age + tidy_data$status)
summary(thalach_age_model)
plot(thalach_age_model)
```

The plot function above is the base R's 4 built-in regression diagnostic plots:

*We will focus on the first two regression diagnostic plots*

* The first plot shows a line that is relatively flat and the relationship between the fitted values and residual is cloud shaped. This implies linearity between the three variables of age, maximum heart rate achieved, and disease status.

* The second plot shows that if the residuals are normally distributed, the points should fall in a diagonal line as it does!

*This, along with the p-values in the summary, show that our linear model effectively captures the relationship between the three variables!*

## Logistic Linear Regression - Binomial and Disease Status

We will now fit a generalized linear regression model to our data to see if we can predict the status of a patient based on the variables within the data set.

```{r}
predicted <- glm(status ~ ., family = "binomial", data = tidy_data)
summary(predicted)
```
As shown by the stars next to the p-values, there are variables that are statistically significant to predict the probability of disease status

* Such as being male, chest pains, etc.

Because we are using a binomial generalized linear model, we are trying to predict the probability of **NOT** getting heart disease due to zero representing heart disease and one representing no heart disease.

* *If the values were flipped and heart disease was represented by one and no heart disease was represented by zero, the binomial generalized linear model would predict the probability of getting heart disease.*

**Below is the scatter plot representing our Binomial Linear Regression Model:**

* To provide context to the below graph, the lower the rank (closer to one) the higher the probability of having heart disease. The higher the rank (further away from one) the higher the probability of **NOT** having heart disease.
```{r}
probability_data <- data.frame(fitted.values = predicted$fitted.values, status = tidy_data$status)

probability_data <- probability_data %>%
  arrange(fitted.values)


probability_data <- probability_data %>%
  mutate(rank = 1:nrow(probability_data))

ggplot(probability_data, aes(x = rank, y = fitted.values, color = status)) +
  geom_point(alpha = 1, shape = 1, stroke = 2) +
  xlab("Rank") +
  ylab("Predicted Probability of Not Getting the Disease")

```

By plotting our predicted values to the known status of someone having the disease or not having the disease, we can confidently say that our generalized linear model predicts the relationship between variables accurately.

*Therefore, given a sample of the same variables, we can predict the likelihood of an individual not getting heart disease depending on their values!*

## Conclusion

WORK ON AFTER REPORT IS FINALIZED


















